Galaxy image statistical generator: how to be confident?

### 2024-09-09 paper/main.tex
more on flows

The ability of generative models have been rapidly adopted in many domains. For instance, in High Energy Physics the reader may be interested by the review \cite{PhysRevD.107.076017}. Concerning Astrophysics and Cosmology the generative models are used in different tasks (e.g. deblending \citep{Hemmati_2022,Arcelin2020} {\color{red} à compléter}) and especially to create synthetic galaxy images of complex morphologies going beyond parametrized analytic light profile simulations (eg. \texttt{GalSim} by \cite{ROWE2015121}). Along this line \citep{ravanbakhsh2016,Fussell2019} use GAN models, \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model (hereafter nicknamed DSDM).

@@

%
\begin{figure}
\resizebox{\columnwidth}{!}{%
%\begin{center}
\begin{tikzpicture}[
node distance = 1cm, 
distrib/.style={rectangle},
]

@@

The structure of the transformation (e.g. Eq.~\ref{eq-NVP_forward}) is a single split of the $\bm{z}$ vector coordinates in two parts: the first part $(z_{\leq s})$ is not transformed while the transformation of the second part ${z_{>s}})$ uses a function depending only on the first part. Notice that the conditioner functions do depends only on $(z_{\leq s})$ for both the forward or the backward transformations. This single split schema is illustrated on Figure \ref{fig-nvp}. 
%
\begin{figure}
\centering
\includegraphics[width=0.45\columnwidth]{fig-nvp.pdf}
\includegraphics[width=0.45\columnwidth]{fig-nvp-inverse.pdf}
\caption{Affine coupling layer schema of forward and backward directions (Eqs.~\ref{eq-NVP_forward},\ref{eq-NVP_backward}) with a single split mechanism (figure inspired by \cite{Papamakarios2021}).}
\label{fig-nvp}
\end{figure}
%
This splitting schema implemented in a so-called \textit{coupling layer} makes the flow-based model computationally efficient and using deep neural networks, one can achieve complex $(\bm{\alpha},\bm{\beta})$ functions. Now, designers of complete architectures can compose different \textit{coupling layers} by reversing $\bm{z}$ elements between layers to ensure that all elements are transformed at the end of the set of operations. An generalisation of any permutation ordering of the inputs can be performed using a $1\times 1$ convolution layer. \cite{2016arXiv160508803D} improves the modelling using a multi-scale approach which would be prohibitive to describe here, and \cite{Kingma2018} use all these extensions to build the \texttt{Glow} model whose schema is displayed on Figure \ref{fig-Glow-archi}.
\begin{figure}
    \centering
    \vspace{-0cm}
    \begin{subfigure}[b]{0.45\textwidth}
    \begin{center}
        \centering
        \includegraphics[width=.7\textwidth]{fig-Glow-1.pdf}
        \caption{One step of Glow flow.}
    \end{center}
    \end{subfigure}%
    \vspace{5mm}
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=.99\textwidth]{fig-Glow-2.pdf}
        \caption{Multi-scale architecture \citep{2016arXiv160508803D}.}
    \end{subfigure}
    \caption{Figure from \citep{Kingma2018}. The flow step is composed of a scale and bias layer with data dependent initialization (\texttt{actnorm}), a $1\times 1$ invertible convolution layer and an affine coupling layer with $(\bm{\alpha},\bm{\beta})$ issued from a shallow convolutional neural network. Both $\bm{x}$ and $\bm{y}$ are tensors of
shape $[h \times w \times c]$ with spatial dimensions $(h, w)$ and channel dimension $c$ ans $(i, j)$ denote
spatial indices into tensors $\bm{x}$ and $\bm{y}$.  This flow step is embedded in a multi-scale  architecture which has $K$ flow steps and a number of levels $L$ \citep{2016arXiv160508803D}. The squeezing operation reduces a $s \times s \times c$ tensor  into a $s/2\times s/2 \times 4c$ tensor which mix spatial and channel components. At each scale, each channel image is divided into $2\times 2$ patches ($s=2$) and each patch is squeezed. }
    \label{fig-Glow-archi}
\end{figure}
 %
\subsection{Score based diffusion models}
%
To generate samples from $p(\bm{x})$, we can proceed as follows \citep[e.g][]{Chang2023,LinYang2023}. Starting from a sample $\bm{x}$ assumed to be drawn from the distribution $p(\bm{x})$, we progressively transform it until the  distribution is as simple as for instance $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.  The idea is then to reverse the process (the \textit{backward} mode) by generating a new sample from a new instance of noise. This transfer mapping looks very similar in concept as normalizing flows. Their latent spaces have the same dimension as the data space, but the main difference consists in the fact that the later are deterministic by essence while the former are stochastic.

The transport process used in the \textit{score diffusion} algorithm is the Ornstein-Uhlenbeck equation \citep{Uhlenbeck1930}. The \textit{forward} process follows the following stochastic differential equation:
\begin{equation}
d\bm{x}_t = -\bm{x}_t dt + \sqrt{2}\ dB_t \quad t\in[0,T]
\label{eq-Ornstein-cont}
\end{equation}
where $dB_t$ is a Wiener process (Brownian motion). The solution of such forward process reads
\begin{equation}
\bm{x}_t = \bm{x}_0 e^{-t} + \underbrace{(1-e^{-2t})^{1/2}}_{\sigma_t} \bm{z} \qquad \bm{z} \overset{iid}{\sim} \mathcal{N}(z,\bm{0},\bm{1})
\end{equation}

>>>>

The ability of generative models have been rapidly adopted in many domains. For instance, in High Energy Physics the reader may be interested by the review \cite{PhysRevD.107.076017}. Concerning Astrophysics and Cosmology the generative models are used in different tasks (e.g. deblending \citep{Hemmati_2022,Arcelin2020} {\color{red} à compléter}) and especially to create synthetic galaxy images of complex morphologies going beyond parametrized analytic light profile simulations (eg. \texttt{GalSim} by \cite{ROWE2015121}). Along this line \citep{ravanbakhsh2016,Fussell2019} use GAN models, \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model.

@@

%
\begin{figure}
\resizebox{0.8\columnwidth}{!}{%
%\begin{center}
\begin{tikzpicture}[
node distance = 1cm, 
distrib/.style={rectangle},
]

@@

The structure of the transformation (e.g. Eq.~\ref{eq-NVP_forward}) is a single split of the $\bm{z}$ vector coordinates in two parts: the first part $(z_{\leq s})$ is not transformed while the transformation of the second part ${z_{>s}})$ uses a function depending only on the first part. Notice that the conditioner functions do depends only on $(z_{\leq s})$ for both the forward or the backward transformations. This single split schema is illustrated on Figure \ref{fig-nvp}. 
%
\begin{figure}
\centering
\includegraphics[width=0.45\columnwidth]{fig-nvp.pdf}
\includegraphics[width=0.45\columnwidth]{fig-nvp-inverse.pdf}
\caption{Affine coupling layer schema of forward and backward directions (Eqs.~\ref{eq-NVP_forward},\ref{eq-NVP_backward}) with a single split mechanism (figure inspired by \cite{Papamakarios2021}).}
\label{fig-nvp}
\end{figure}
%
This splitting schema implemented in a so-called \textit{coupling layer} makes the flow-based model computationally efficient and using deep neural networks, one can achieve complex $(\bm{\alpha},\bm{\beta})$ functions. Now, designers of complete architectures can compose different \textit{coupling layers} by reversing $\bm{z}$ elements between layers to ensure that all elements are transformed at the end of the set of operations. An generalisation of any permutation ordering of the inputs can be performed using a $1\times 1$ convolution layer. \cite{2016arXiv160508803D} improves the modelling using a multi-scale approach which would be prohibitive to describe here, and \cite{Kingma2018} use all these extensions to build the \texttt{Glow} model whose schema is displayed on Figure \ref{fig-Glow-archi}.
\begin{figure}
    \centering
    \vspace{-0cm}
    \begin{subfigure}[b]{0.45\textwidth}
    \begin{center}
        \centering
        \includegraphics[width=.7\textwidth]{fig-Glow-1.pdf}
        \caption{One step of Glow flow.}
    \end{center}
    \end{subfigure}%
    \vspace{5mm}
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=.99\textwidth]{fig-Glow-2.pdf}
        \caption{Multi-scale architecture \citep{2016arXiv160508803D}.}
    \end{subfigure}
    \caption{Figure from \citep{Kingma2018}. The flow step is composed of a scale and bias layer with data dependent initialization (\texttt{actnorm}), a $1\times 1$ invertible convolution layer and an affine coupling layer with $(\bm{\alpha},\bm{\beta})$ issued from a shallow convolutional neural network. Both $\bm{x}$ and $\bm{y}$ are tensors of
shape $[h \times w \times c]$ with spatial dimensions $(h, w)$ and channel dimension $c$ ans $(i, j)$ denote
spatial indices into tensors $\bm{x}$ and $\bm{y}$.  This flow step is embedded in a multi-scale  architecture which has $K$ flow steps and a number of levels $L$ \citep{2016arXiv160508803D}. The squeezing operation reduces a $s \times s \times c$ tensor  into a $s/2\times s/2 \times 4c$ tensor which mix spatial and channel components. At each scale, each channel image is divided into $2\times 2$ patches ($s=2$) and each patch is squeezed. }
    \label{fig-Glow-archi}
\end{figure}
 %
\subsection{Score based diffusion models}
%
To generate samples from $p(\bm{x})$, we can proceed as follows \citep[e.g][]{Chang2023,LinYang2023}. Starting from a sample $\bm{x}_0$ assumed to be drawn from the distribution $p_0(\bm{x})=p(\bm{x})$, we progressively transform it until the  distribution is as simple as for instance $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.  The idea is then to reverse the process (the \textit{backward} mode) by generating a new sample from a new instance of noise. This transfer mapping looks very similar in concept as normalizing flows. Their latent spaces have the same dimension as the data space, but the main difference consists in the fact that the later are deterministic by essence while the former are stochastic.

Following the notations used in \cite{kadkhodaie2024generalization}, the transport process used in the \textit{score diffusion} algorithm is the Ornstein-Uhlenbeck equation \citep{Uhlenbeck1930}. The \textit{forward} process follows the following stochastic differential equation:
\begin{equation}
d\bm{x}_t = -\bm{x}_t dt + \sqrt{2}\ dB_t \quad t\in[0,T]
\label{eq-Ornstein-cont}
\end{equation}
where $dB_t$ is a Wiener process (Brownian motion) and $T$ is a finite ending time that would be considered to be large enough. The solution of such forward process reads
\begin{equation}
\bm{x}_t = \bm{x}_0 e^{-t} + \underbrace{(1-e^{-2t})^{1/2}}_{\sigma_t} \bm{z} \qquad \bm{z} \overset{iid}{\sim} \mathcal{N}(\bm{z},\bm{0},\bm{1})
\end{equation}
The associated probability density function is given by the expression 
\begin{equation}
p_t(\bm{x}_t) = \int p_t(\bm{x}_t,\bm{x}_0) d\bm{x}_0 = \int p_t(\bm{x}_t|\bm{x}_0) p_0(\bm{x}_0)dx_0
\label{eq-smoothing-p0-gauss}
\end{equation}
where according to the Fokker-Planck representation of the Ornstein-Uhlenbeck process,  it yields
\begin{equation}
p_t(\bm{x}_t|\bm{x}_0)= \frac{1}{(2\pi \sigma_t^2)^{d/2}}\exp\left\{\displaystyle -\frac{\|\bm{x}_t-\bm{x}_0e^{-t}\|^2}{2\sigma_t^2}\right\}
\end{equation}
So, $p_t(\bm{x}_t)$ is the result of the convolution of the initial distribution $p_0(x_0)$ with a Gaussian with variance $\sigma_t^2$ which gradually tends towards the normal distribution $\mathcal{N}(.,\bm{0},\bm{1})$, producing a progressive blurring that also makes $p_t(\bm{x}_t)$ converging towards the normal distribution.

Notice that in a VAE architecture beyond the information compression,  the transformation from $\bm{x}_0$ to $\bm{z}$ is performed by an \textit{encoder} whose implementation is typically a deep neural networks with trained parameters. This is not the case in the above stochastic process where only noise is added to the original signal (e.g. galaxy image). In the same spirit, the VAE \textit{decoder} or the inverted flows are replaces by stochastic differential equation (damped Langevin)
\begin{equation}
dx_{T-t} = (x_{T-t} + 2\nabla_x \log p_{T-t}(x_{T-t}))dt + \sqrt{2}\ dB_t \quad t\in[0,T]
\label{eq-backward-diffusion}
\end{equation}
where $\nabla_x \log p_{t}(\bm{x}_{t})$ is the \textit{score} of the probability density at time $t$. This backward process is the $\bm{x}=\bm{x}_0$ generative direction from $\mathcal{N}(\bm{z},\bm{0},\bm{1})$ sampling.

The key problem is how to obtain the score. One can think about two possible approaches: either we have a model for $p_t$ \citep[e.g.][]{Guth2022b,Lempereur2024} that can exploit regularity patterns and assumptions about correlations to provide a constrained parametric framework for learning the score by score matching \citep{hyvarinen2005a}, or we use a \textit{denoising} neural network as introduced by \cite{Bengio2013} and then \cite{Sohl-Dickstein2015,Ho2020} and recently studied from a mathematical point of view by \citep{kadkhodaie2024generalization}, as well as used in cosmology such as the generation of galaxy images \citep{smith2021}, the generation of 21cm luminosity temperature maps \citep{Zhao2023} and super-resolved large-scale cosmic structure predictions \citep{Schanz2023} and sampling of the high-dimensional Bayesian posterior of the weak lensing mass-mapping problem \citep{Remy2023}.

Denoising diffusion stochastic models have certain advantages compared to GAN....
